### 2014-12-16 text/Chapter06.tex
Spell checking

\subsection{Shapes and containers}
Our generalization of coeffect structure using a \emph{coeffect container} is based on containers of 
Abbott et al. \cite{types-containers}. Interestingly, containers have later been linked
to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of 
lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise 
as some containers may have multiple shapes of the same length, \eg~differently ballanced trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

@@

\item The contianer representing the singleton data type is given by $S \triangleleft P$ where
  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
  shape $\ast$ contains exactly one position. To follow the intuition based on offset or 
  index, we write the single position as $0$, that is $P(\ast) = \{ 0 \}$.
\end{itemize}

@@

\noindent
As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
an underlying category structre; $\cunit$ and $\czero$ represent an accessed and unused variable, 
respectively.

@@

The elements $\sempty$ and $\sunit$ represent the shapes of the empty and the singleton free-variable 
context, respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts.
Given $\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$. 
\end{definition}

@@

The type system for the unified coeffect calculus is parameterized by a coeffect scalar 
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$ 
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$. 
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
use of variables -- the associativity is built-in. The order of variables matters, but can be changed 
using the exchane rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ 
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars.

@@

In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
system is that the vector concatentation $\atimes$ is now replaced with context splitting/merging 
of the unified coeffect algebra. As already mentioned, we use splitting of context requirements $\azip$
in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.

@@

\noindent
In the last two cases, we need to turn the coeffect into a form that is required by the exchange
and contraction rules. Aside from idempotence, we could use the unit property and obtain 
\eg~$\czero\,\czip\,\cclrd{r}\,\czip\,\czero\,\czip\,\czero$. However, this appraoch does not 
work because $\cpar$ and $\czip$ may have different unit elements (in fact, we do not ever require
the existence of a unit for $\czip$).

@@

\subsection{Semanitcs of unified calculus}
\label{sec:unified-semantics}

@@

The semantics of the unified coeffect calculus follows the same pattern as the semantics of the
structural calculus (Section~\ref{sec:struct-semantics}) with a number of differences. The 
index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of 
the structure $\C^{\SP(\sclrd{n})}$ for a specified container $\SH \triangleleft \SP$ and a shape 
$\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in 
Section~\ref{sec:unified-types}, we need to modify the domains and codomains of the \ident{merge}, 
\ident{split} and \ident{dup} operations as follows.
%
\begin{equation*}
\begin{array}{lcl}
 \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}& 
  \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \\
 \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}& 
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \\
 \ident{dup}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
  \ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}\,\azip\,\alift{\cclrd{s}}}{(\alpha \xtimes \alpha)} \\
\end{array}
\end{equation*}
%
The key difference is that the vector concatenation $\atimes$ is replaced with $\apar$ in the
\emph{domain} of the operations (in \ident{split}) and with $\azip$ in the codomain
(in \ident{merge}, but also in \ident{dup} which models contraction).
This means that the splitting and merging the context can change the coeffect annotation:
%
\begin{equation*}
\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~:~
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
  \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} 
\end{equation*}
%
For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
splitting and then merging a context may lose some of the avilable contextual capabilities
(and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:unified-rev-contr} 
and Lemma~\ref{thm:unified-rev-exch} relied on sub-coeffecting in the flat system.

@@

More lightweight syntax for effectful computation can be obtained using
techniques that automatically insert the necessary monadic plumbing (without
special syntax). This has been done for effecful computations \cite{monads-lightweight-ml}
and a similar approach would be worth exploring for coeffects.

@@

The other aspect of practical implementation of coeffects is that of tracking the context requirements
(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
language feature) the type system needs to be able to capture various kinds of coeffect algebras. 
The structures required in this thesis include sets (with union or interesection), natural numbers (with 
addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids 
(vectors of annotations).

@@

Compared to typical effect systems based on sets, coeffect systems have richer structure consisting
of four operations ($\cseq,\czip,\cpar$ and a relation $\cleq$). This allows capturing interesting 
properties, but it makes working with coeffect annotations difficult -- it compilcates tasks
such as type generalization and constraint solving in type inference. Here, we look at one possible 
simplification of the structure.

@@

\noindent
The definition provides all the operations of a flat coeffect algebra, but it defines a more rigid 
structure. In particular, $\czip$ and $\cpar$ are both defined in terms of $\cleq$. This simplifies 
the system as the operations of the underlying indexed comonad structure can be expressed using just
$\cseq$ and $\cleq$:
%
\begin{equation*}
\begin{array}{lcll}
 \ident{counit}_{\cunit} &\narrow{:}& 
    \ctyp{\cunit}{\alpha} \rightarrow \alpha \\
 \ident{cobind}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
    (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) \\
\ident{merge}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}& 
    \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{t}}{(\alpha \times \beta)} &
    (\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s}) \\
\ident{split}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}& 
    \ctyp{\cclrd{t}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} &
    (\cclrd{r}\,\cleq\,\cclrd{t}, \cclrd{s}\,\cleq\,\cclrd{t}) \\
\ident{lift}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
    \ctyp{\cclrd{r}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\alpha} & 
    (\cclrd{s}\,\cleq\,\cclrd{r})
\end{array}
\end{equation*}
%
Note that we still need \emph{two distinct} operators in the system. Analysis of our examples shows
that the operator used for sequential composition $\cseq$ is not necessarilly related in any way to 
the ordering on the lattice provided by $\cleq$.

@@

\begin{figure}[t]
{\small a.) Littice-based flat systems (selected rules):}

@@

\vspace{1.5em}
{\small b.) Littice-based structural systems (selected rules):}

@@

In smmary, the lattice-based coeffect system has some advantages over the design used in this
thesis. In particular, it uses better-known structures and is simpler. However, we do not use
it as it rules out one of the important motivating examples (implicit parameters) and, even with
the simplification, solving constraints with $\cseq$ and $\cleq$ is still an open question.

@@

The coeffect meta-langauge could be designed using both flat and structural indexed comonads.
For simplicity, this section only discusses the flat variant. The syntax of types and terms
of the langauge includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four additional
language constructs:
%
\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
 e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 
\end{array}
\end{equation*}
%
The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
operation of the comonad. To define meta-langauge for flat indexed comonads, we also include 
\kvd{split} and \kvd{merge} that embed the corresponding operations.

@@

Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in the flat coeffect calculus,
we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its emebdding in the coeffect
meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
to represent the entire variable context of the source language. The translation is defined in
Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in 
Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
the operations of an indexed comonad.

@@

In the rest of the chapter, we discussed two topis of future and related work. First
(Section~\ref{sec:unified-impl}), we considered pathways to practical implementations of
coeffect systems, including a discussion of a novel lattice-based coeffect system, which is simpler,
but cannot capture all our motivating examples. Finally, we discussed how our work relates to 
meta-language based on comonads (Section~\ref{sec:unified-meta}). We present a \emph{coeffect 
meta-language} that follows similar style to CMTT, but is based on indexed comonads. Next, we 
discussed how to embed flat coeffect systems in this coeffect meta-language, which elucidates the 
relationship between our work and CMTT.

>>>>

\subsection{Shapes and containers}
Our generalization of coeffect structure using a \emph{coeffect container} is based on containers of 
Abbott et al. \cite{types-containers}. Interestingly, containers have later been linked
to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of 
lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise 
as some containers may have multiple shapes of the same length, \eg~differently balanced trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

@@

\item The container representing the singleton data type is given by $S \triangleleft P$ where
  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
  shape $\ast$ contains exactly one position. To follow the intuition based on offset or 
  index, we write the single position as $0$, that is $P(\ast) = \{ 0 \}$.
\end{itemize}

@@

\noindent
As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
an underlying category structure; $\cunit$ and $\czero$ represent an accessed and unused variable, 
respectively.

@@

The elements $\sempty$ and $\sunit$ represent the shapes of the empty and the singleton free-variable 
context, respectively. The $\stimes$ operation corresponds to concatenation of free-variable contexts.
Given $\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$. 
\end{definition}

@@

The type system for the unified coeffect calculus is parameterized by a coeffect scalar 
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$ 
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$. 
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
use of variables -- the associativity is built-in. The order of variables matters, but can be changed 
using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ 
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars.

@@

In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
system is that the vector concatenation $\atimes$ is now replaced with context splitting/merging 
of the unified coeffect algebra. As already mentioned, we use splitting of context requirements $\azip$
in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.

@@

\noindent
In the last two cases, we need to turn the coeffect into a form that is required by the exchange
and contraction rules. Aside from idempotence, we could use the unit property and obtain 
\eg~$\czero\,\czip\,\cclrd{r}\,\czip\,\czero\,\czip\,\czero$. However, this approach does not 
work because $\cpar$ and $\czip$ may have different unit elements (in fact, we do not ever require
the existence of a unit for $\czip$).

@@

\subsection{Semantics of unified calculus}
\label{sec:unified-semantics}

@@

The semantics of the unified coeffect calculus follows the same pattern as the semantics of the
structural calculus (Section~\ref{sec:struct-semantics}) with a number of differences. The 
index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of 
the structure $\C^{\SP(\sclrd{n})}$ for a specified container $\SH \triangleleft \SP$ and a shape 
$\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in 
Section~\ref{sec:unified-types}, we need to modify the domains and codomains of the \ident{merge}, 
\ident{split} and \ident{dup} operations as follows.
%
\begin{equation*}
\begin{array}{lcl}
 \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}& 
  \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \\
 \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}& 
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \\
 \ident{dup}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
  \ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}\,\azip\,\alift{\cclrd{s}}}{(\alpha \xtimes \alpha)} \\
\end{array}
\end{equation*}
%
The key difference is that the vector concatenation $\atimes$ is replaced with $\apar$ in the
\emph{domain} of the operations (in \ident{split}) and with $\azip$ in the codomain
(in \ident{merge}, but also in \ident{dup} which models contraction).
This means that the splitting and merging the context can change the coeffect annotation:
%
\begin{equation*}
\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~:~
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
  \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} 
\end{equation*}
%
For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
splitting and then merging a context may lose some of the available contextual capabilities
(and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:unified-rev-contr} 
and Lemma~\ref{thm:unified-rev-exch} relied on sub-coeffecting in the flat system.

@@

More lightweight syntax for effectful computation can be obtained using
techniques that automatically insert the necessary monadic plumbing (without
special syntax). This has been done for effectful computations \cite{monads-lightweight-ml}
and a similar approach would be worth exploring for coeffects.

@@

The other aspect of practical implementation of coeffects is that of tracking the context requirements
(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
language feature) the type system needs to be able to capture various kinds of coeffect algebras. 
The structures required in this thesis include sets (with union or intersection), natural numbers (with 
addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids 
(vectors of annotations).

@@

Compared to typical effect systems based on sets, coeffect systems have richer structure consisting
of four operations ($\cseq,\czip,\cpar$ and a relation $\cleq$). This allows capturing interesting 
properties, but it makes working with coeffect annotations difficult -- it complicates tasks
such as type generalization and constraint solving in type inference. Here, we look at one possible 
simplification of the structure.

@@

\noindent
The definition provides all the operations of a flat coeffect algebra, but it defines a more rigid 
structure. In particular, $\czip$ and $\cpar$ are both defined in terms of $\cleq$. This simplifies 
the system as the operations of the underlying indexed comonad structure can be expressed using just
$\cseq$ and $\cleq$:
%
\begin{equation*}
\begin{array}{lcll}
 \ident{counit}_{\cunit} &\narrow{:}& 
    \ctyp{\cunit}{\alpha} \rightarrow \alpha \\
 \ident{cobind}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
    (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) \\
\ident{merge}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}& 
    \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{t}}{(\alpha \times \beta)} &
    (\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s}) \\
\ident{split}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}& 
    \ctyp{\cclrd{t}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} &
    (\cclrd{r}\,\cleq\,\cclrd{t}, \cclrd{s}\,\cleq\,\cclrd{t}) \\
\ident{lift}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
    \ctyp{\cclrd{r}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\alpha} & 
    (\cclrd{s}\,\cleq\,\cclrd{r})
\end{array}
\end{equation*}
%
Note that we still need \emph{two distinct} operators in the system. Analysis of our examples shows
that the operator used for sequential composition $\cseq$ is not necessarily related in any way to 
the ordering on the lattice provided by $\cleq$.

@@

\begin{figure}[t]
{\small a.) Lattice-based flat systems (selected rules):}

@@

\vspace{1.5em}
{\small b.) Lattice-based structural systems (selected rules):}

@@

In summary, the lattice-based coeffect system has some advantages over the design used in this
thesis. In particular, it uses better-known structures and is simpler. However, we do not use
it as it rules out one of the important motivating examples (implicit parameters) and, even with
the simplification, solving constraints with $\cseq$ and $\cleq$ is still an open question.

@@

The coeffect meta-language could be designed using both flat and structural indexed comonads.
For simplicity, this section only discusses the flat variant. The syntax of types and terms
of the language includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four additional
language constructs:
%
\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
 e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 
\end{array}
\end{equation*}
%
The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
operation of the comonad. To define meta-language for flat indexed comonads, we also include 
\kvd{split} and \kvd{merge} that embed the corresponding operations.

@@

Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in the flat coeffect calculus,
we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its embedding in the coeffect
meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
to represent the entire variable context of the source language. The translation is defined in
Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in 
Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
the operations of an indexed comonad.

@@

In the rest of the chapter, we discussed two topics of future and related work. First
(Section~\ref{sec:unified-impl}), we considered pathways to practical implementations of
coeffect systems, including a discussion of a novel lattice-based coeffect system, which is simpler,
but cannot capture all our motivating examples. Finally, we discussed how our work relates to 
meta-language based on comonads (Section~\ref{sec:unified-meta}). We present a \emph{coeffect 
meta-language} that follows similar style to CMTT, but is based on indexed comonads. Next, we 
discussed how to embed flat coeffect systems in this coeffect meta-language, which elucidates the 
relationship between our work and CMTT.

### 2014-12-16 text/Chapter07.tex
Spell checking

Coeffect annotations in context-aware programming languages can contain significant amount
of information. Thus, the programming langauge also needs a form of type inference that can propagate
such information. This has been done for effect systems \cite{effects-polymorphic}. For 
\emph{flat coeffects}, type inference is complicated by the non-determinism in the typing rule
for lambda abstraction, but the problem is simpler for \emph{structural coeffects}. In addition
to the \emph{algebra providers} (discussed in Section~\ref{sec:unified-impl-types}), it would also
be worth developing a bidirectional type system \cite{types-bidirectional} for coeffects, where 
\emph{some} annotations are required, but most types and coeffects are reconstructed automatically.

@@

In the \emph{flat} and \emph{sturctural} coeffect calculi, we attach annotations to the \emph{entire
context} and to \emph{individual variables}, respectively. We later unified the two systems 
in Section~\ref{sec:unified-unified}.

@@

\paragraph{Applications.}
The last direction for further development is developing practical programming langauges or
libraries based on the theory presented in this thesis. We believe that the most useful approach
is to extend programming languages with a reusable coeffect system and allow library developers 
define their own coeffect algebras. This option has been discussed in Section~\ref{sec:unified-impl}.
For languages like Haskell, this can be done through a light-weight syntactic sugar akin to the
``do'' notation.

>>>>

Coeffect annotations in context-aware programming languages can contain significant amount
of information. Thus, the programming language also needs a form of type inference that can propagate
such information. This has been done for effect systems \cite{effects-polymorphic}. For 
\emph{flat coeffects}, type inference is complicated by the non-determinism in the typing rule
for lambda abstraction, but the problem is simpler for \emph{structural coeffects}. In addition
to the \emph{algebra providers} (discussed in Section~\ref{sec:unified-impl-types}), it would also
be worth developing a bidirectional type system \cite{types-bidirectional} for coeffects, where 
\emph{some} annotations are required, but most types and coeffects are reconstructed automatically.

@@

In the \emph{flat} and \emph{structural} coeffect calculi, we attach annotations to the \emph{entire
context} and to \emph{individual variables}, respectively. We later unified the two systems 
in Section~\ref{sec:unified-unified}.

@@

\paragraph{Applications.}
The last direction for further development is developing practical programming languages or
libraries based on the theory presented in this thesis. We believe that the most useful approach
is to extend programming languages with a reusable coeffect system and allow library developers 
define their own coeffect algebras. This option has been discussed in Section~\ref{sec:unified-impl}.
For languages like Haskell, this can be done through a light-weight syntactic sugar akin to the
``do'' notation.
